Results

### 2022-08-02 article/intro.Rmd
most meth rev MS

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.

@@

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

>>>>

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**

@@

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
**- what is new in this study?**
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

### 2022-08-02 article/merge.Rmd
most meth rev MS

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author: 
  - Nicholas Medina ^a^*
  - Max Villalobos ^b^
  - Andrea Johnson ^b^
date: 
  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

@@

<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->

@@

Thanks to Professor Dan Perlman for initial networking; 
<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; 
peer James Fifer for discussion of early draft; 
and Luisa Valdez for translation assistance.

>>>>

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author:
  - Nicholas Medina * ^a^
  # - Emma Capetz ^c^
  - Andrea Johnson ^b^
  # - Erin Lau ^c^
  # - Agustín Mendoza ^b^
  # - Don Morris ^c^
  - Max Villalobos ^b^
date: \scriptsize
  ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
  # ^a^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

@@

<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), Small-scale Forestry (1.3), Eco Evo -->

@@

Thanks to 
Professor Dan Perlman for initial networking;
<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
James Fifer, Chau Ho, and Michelle Spicer for discussion of early drafts;
and Luisa Valdez for translation assistance.

### 2022-08-02 article/merge.tex
most meth rev MS

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{}
\date{\vspace{-2.5em}}

@@

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/map-1.pdf}
\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2009 Google, © 2022 Osa Conservation. Photo credit: Nicholas Medina.}
\end{figure}

@@

--\textgreater{}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina * a; Andrea Johnson b; Max Villalobos b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina * \textsuperscript{a} \and Andrea Johnson \textsuperscript{b} \and Max Villalobos \textsuperscript{b}}
\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}

@@

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/map-1.pdf}
\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation, © 2009 Google. Map GIS data and photo credit: Max Villalobos, Nicholas Medina.}
\end{figure}

@@

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

Thanks to
Professor Dan Perlman for initial networking;
--\textgreater{}
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
peer James Fifer for discussion of early draft;
and Luisa Valdez for translation assistance.

\hypertarget{author-contributions}{%
\section*{Author contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author contributions}

MV and AJ contributed to study conceptualization, administration, and supervision;
NM and AJ contributed to funding acquisition;
NM contributed to data collection;
NM and MV contributed to data analysis;
NM wrote initial draft and later revisions.

### 2022-08-02 article/methods.Rmd
most meth rev MS

```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2009 Google, © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

@@

The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,000 trees.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
<!-- *[References used for this step are given in *Table S1*.] * -->
Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations. 
<!-- However, these updates did not appear to change the main results of this study. -->

Traits including successional stage and main dispersal mode were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.

@@

For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
 <!-- including biomass converted to a per hectare basis,  -->

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.

>>>>

```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation, © 2009 Google. Map GIS data and photo credit: Max Villalobos, Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```

@@

Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
Given the small size of the last 300 m distance stratum (Fig \@ref(fig:map)c), one census plot was sufficiently representative.
All data were later centered at the stratum level, so resulting trends were not weighted by distance bands with more plots in them. 
The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

@@

The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,243 trees.
<!-- Following common forestry guidelines,  -->
In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified by local experts, 
and trait information was gathered from the literature.
Traits including successional stage (early, late) and main dispersal mode (wind, animal, water) were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.

@@

Aboveground biomass (AGB) was calculated at the individual stem level using the allometric equation by *@chave15* and later centered at the plot level.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
<!-- Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org). -->
<!-- *[References used for this step are given in *Table S1*.] * -->
Values were updated using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations. 
<!-- However, these updates did not appear to change the main results of this study. -->

@@

For all response variables, data were centered using medians at the plot level, followed by the discrete distance stratum level where applicable, which was most cases and indicated on figures. 
These medians were then subjected to linear regression with distance to intact forest edge as the only independent variable, which was binned discretely according to census design.
All regressions were run through assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
 <!-- including biomass converted to a per hectare basis,  -->

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis, including unidentified taxa, which were grouped together for community analyses and did not affect results.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.

### 2022-08-02 article/results.Rmd
most meth rev MS

```{r setupRes, echo = F, message = F, warning=F, results='hide'}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
source(here::here("figs/fig2.R")); here::i_am("article/results.Rmd"); 
options(digits = 1)
# source(here::here("figs/fig2.R"));
```

@@

```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa. White and colored dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig2.R"), echo = F);
fig2
# knitr::include_graphics(here::here("figs/fig2.png"))
```

@@

```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig3.R"), echo = F);
fig3
# knitr::include_graphics(here::here("figs/fig3.png"))
```

>>>>

```{r setupRes, echo = F, message = F, warning=F, results='hide'}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
source(here::here("figs/fig2.R")); here::i_am("article/results.Rmd"); 
options(digits = 2)
# source(here::here("figs/fig2.R"));
```

@@

```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa. White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig2.R"), echo = F);
fig2
# knitr::include_graphics(here::here("figs/fig2.png"))
```

@@

```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig3.R"), echo = F);
fig3
# knitr::include_graphics(here::here("figs/fig3.png"))
```
